Modeling Outbreak Frequency by Season and Pathogen: Insights and Predictions for 2025

### 2024-12-03 paper/paper.qmd
Update the data source and citation

The data for this study was sourced from Open Data Toronto and downloaded using the @citeopendatapkg package. It consists of 846 observations from 2024 across 10 variables.

Our analysis was conducted using the statistical programming language R [@citeR], leveraging a suite of packages to facilitate efficient data manipulation, modeling, and visualization. Data cleaning, transformation, and visualization were handled using `tidyverse` @tidyverse and `dplyr` @dplyr. The core analysis used Poisson regression models to explore relationships between outbreak counts and key variables, including pathogen type, seasonal variation, and outbreak settings. These models were implemented using the flexible framework offered by `rstanarm` @rstanarm for model development and extensions. Efficient data integration was achieved through `arrow` @arrow, and modeling workflows were supported using `modelr` @modelr within the tidyverse framework. Model summaries were generated using `modelsummary` @modelsummary, while `ggplot2` @ggplot2 was employed for creating visualizations. Enhanced table outputs were produced with `kableExtra` @kableExtra, and dynamic report generation was performed with `knitr` @citeknitr. File management, including reading parquet files, was streamlined with `here` @citehere.

The analysis framework followed the structured approach provided by @citeRohan, ensuring consistency and clarity in integrating data preparation and modeling workflows. Additionally, the critique and guidelines outlined in @tellingstories informed our presentation of the data and results, emphasizing clarity and coherence.

@@

The cleaning process for the dataset focused on retaining key variables relevant to respiratory outbreaks in Toronto healthcare institutions. Initially, redundant or irrelevant columns were removed, and the dataset was narrowed to include only `institution_address`, `outbreak_setting`, `type_of_outbreak`, `causative_agent_1`, `date_outbreak_began`, and `active` variables. Outbreaks were filtered to focus on respiratory pathogens of interest, such as "COVID-19", "Coronavirus*", "Rhinovirus", and "Parainfluenza". The season variable was derived from the `date_outbreak_began` column by mapping months to their respective seasons like December-February as Winter. Invalid or missing data were addressed by removing incomplete rows or invalid entries, ensuring data consistency. These steps ensured that the dataset was streamlined for meaningful analysis while retaining critical details.

The creation of the `count_data` dataset involved aggregating the cleaned data to prepare it for statistical modeling and visualization. Outbreak events were grouped by `causative_agent_1`, `season`, `month`, and `outbreak_setting`, and the `outbreak_count` variable was calculated as the total number of outbreaks for each combination of these factors. This aggregation provided a structured summary of outbreak frequency, enabling comparisons between pathogens, seasonal trends, and healthcare settings. By summarizing the data, `count_data` served as the foundation for constructing Poisson regression models and generating meaningful visualizations to explore outbreak dynamics.

@@

The data used in this study captures respiratory pathogen outbreaks in Toronto healthcare settings, collected through systematic reporting and published via @TorontoOutbreaks2024. Each outbreak is recorded as a discrete event, including the causative pathogen, the setting like hospital, long-term care facility, and key dates: when the outbreak began and when it was declared.

@@

The findings revealed marked seasonal sensitivity for pathogens like Rhinovirus and Parainfluenza, which showed distinct peaks in spring and late summer to early fall, respectively. In contrast, COVID-19 displayed a near-steady transmission rate across the year, reflecting a lack of climatic influence. This consistency aligns with guidance from the World Health Organization, which notes that COVID-19 transmission is driven by human-to-human contact rather than environmental factors @citewho. Rhinovirus and Parainfluenza, however, exhibited seasonality driven by factors such as weather transitions, increased indoor activity, and environmental stability favorable to these pathogens.

@@

The dataset utilized in this study centers on outbreak data collected from healthcare institutions in Toronto, following stringent regulatory guidelines. The data collection is governed by the Ontario Health Protection and Promotion Act (HPPA) @OntarioHealthAct1990, which mandates healthcare institutions—including hospitals, long-term care homes, and retirement homes—to monitor and report signs of respiratory and gastroenteric infections among staff and residents. An outbreak is defined as an unusual increase in infection rates beyond the expected baseline within a specific institution or ward. Reporting is carried out by the institutions to Toronto Public Health (TPH), which investigates the reported outbreaks and collaborates with institutions to implement control measures @TorontoOutbreaks2024.

@@

The dataset includes all reported outbreaks from Toronto's healthcare institutions, representing a near-complete population of healthcare-associated outbreaks. The framework, defined by HPPA requirements, ensures comprehensive coverage of outbreaks within institutions mandated to report. The data essentially serves as a full census of reported outbreaks rather than a traditional sample, as all institutions are legally obligated to report under HPPA. However, certain cases are excluded, such as mild infections managed outside healthcare settings or individuals unable to seek timely medical care, resulting in potential underrepresentation.

@@

Outbreaks are identified and reported by healthcare institutions based on clinical symptoms or laboratory-confirmed cases. These reports are then submitted to Toronto Public Health, with updates provided weekly to maintain real-time relevance @OntarioMinistry2024. However, smaller outbreaks or those diagnosed late may be underreported, leading to data gaps. Variability in reporting practices across institution types further complicates data consistency; for instance, long-term care homes may report more frequently due to heightened surveillance requirements, whereas transitional care centers might report less rigorously.

@@

The dataset adheres to rigorous reporting standards mandated by the HPPA @OntarioHealthAct1990, ensuring uniformity in data collection across institutions. Outbreaks are defined using standardized criteria, such as an increase in infection rates above expected baselines within a specific ward or institution. This ensures consistency in reported data. Institutions work closely with Toronto Public Health (TPH) to verify and classify outbreaks, further enhancing the dataset's reliability.

@@

The dataset supports broader public health efforts to manage and mitigate healthcare-associated infections, aligning with the objectives outlined in the 2023-2024 Canadian Public Health Plan @Canada2024. The outbreak monitoring framework is consistent with international practices, such as the World Health Organization's guidelines for disease surveillance and transmission @citewho. By linking local data collection to global health standards, this dataset highlights the critical role of systematic outbreak reporting in improving public health outcomes and guiding evidence-based policy decisions.

>>>>

Our analysis was conducted using the statistical programming language R [@citeR], leveraging a suite of packages to facilitate efficient data manipulation, modeling, and visualization. Data cleaning, transformation, and visualization were handled using `tidyverse` [@tidyverse] and `dplyr` [@dplyr]. The core analysis used Poisson regression models to explore relationships between outbreak counts and key variables, including pathogen type, seasonal variation, and outbreak settings. These models were implemented using the flexible framework offered by `rstanarm` [@rstanarm] for model development and extensions. Efficient data integration was achieved through `arrow` [@arrow], and modeling workflows were supported using `modelr` [@modelr] within the tidyverse framework. Model summaries were generated using `modelsummary` [@modelsummary], while `ggplot2` [@ggplot2] was employed for creating visualizations. Enhanced table outputs were produced with `kableExtra` [@kableExtra], and dynamic report generation was performed with `knitr` [@citeknitr]. File management, including reading parquet files, was streamlined with `here` [@citehere].

The analysis framework followed the structured approach provided by Rohan [@citeRohan], ensuring consistency and clarity in integrating data preparation and modeling workflows. Additionally, the critique and guidelines outlined in Telling Stories With Data [@tellingstories] informed our presentation of the data and results, emphasizing clarity and coherence.

## Data Source

The dataset used in this study was obtained from Open Data Toronto [@torontooutbreaks], a publicly accessible platform providing high-quality, open government data. It was downloaded using the `opendatatoronto` [@citeopendatapkg], facilitating seamless access to the latest updates. The dataset includes 846 observations from 2024, covering 10 variables related to outbreaks in Toronto's healthcare institutions. These variables encompass key details such as outbreak settings, causative agents, dates of outbreak initiation, and other contextual factors critical for understanding infection patterns. Regular updates ensure the dataset reflects real-time trends, making it a robust resource for public health research and analysis.

@@

The dataset was cleaned to retain variables critical to respiratory outbreaks in Toronto healthcare institutions, including `institution_address`, `outbreak_setting`, `type_of_outbreak`, `causative_agent_1`, `date_outbreak_began`, and `active`. Focus was placed on respiratory pathogens of interest: "COVID-19", "Coronavirus*", "Rhinovirus", and "Parainfluenza". A `season` variable was created by mapping months to their respective seasons, and incomplete or invalid rows were removed to ensure consistency. These steps streamlined the dataset for analysis while preserving essential details.

The count_data dataset was created by aggregating the cleaned data, grouping outbreaks by `causative_agent_1`, `season`, `month`, and `outbreak_setting`, and calculating `outbreak_count` as the total for each group. This structured summary enabled comparisons of pathogen-specific trends, seasonal patterns, and outbreak settings, forming the basis for Poisson regression models and visualizations to explore outbreak dynamics.

@@

The data used in this study captures respiratory pathogen outbreaks in Toronto healthcare settings, collected through systematic reporting and published via Open Data Toronto [@torontooutbreaks]. Each outbreak is recorded as a discrete event, including the causative pathogen, the setting like hospital, long-term care facility, and key dates: when the outbreak began and when it was declared.

@@

The findings revealed marked seasonal sensitivity for pathogens like Rhinovirus and Parainfluenza, which showed distinct peaks in spring and late summer to early fall, respectively. In contrast, COVID-19 displayed a near-steady transmission rate across the year, reflecting a lack of climatic influence. This consistency aligns with guidance from the World Health Organization, which notes that COVID-19 transmission is driven by human-to-human contact rather than environmental factors [@citewho]. Rhinovirus and Parainfluenza, however, exhibited seasonality driven by factors such as weather transitions, increased indoor activity, and environmental stability favorable to these pathogens.

@@

The dataset utilized in this study centers on outbreak data collected from healthcare institutions in Toronto, following stringent regulatory guidelines. The data collection is governed by the Ontario Health Protection and Promotion Act (HPPA) [@OntarioHealthAct1990], which mandates healthcare institutions—including hospitals, long-term care homes, and retirement homes—to monitor and report signs of respiratory and gastroenteric infections among staff and residents. An outbreak is defined as an unusual increase in infection rates beyond the expected baseline within a specific institution or ward. Reporting is carried out by the institutions to Toronto Public Health (TPH), which investigates the reported outbreaks and collaborates with institutions to implement control measures [@TorontoOutbreaks2024].

@@

The dataset includes all reported outbreaks from Toronto's healthcare institutions, representing a near-complete population of healthcare-associated outbreaks. The framework, defined by HPPA requirements, ensures comprehensive coverage of outbreaks within institutions mandated to report. The data essentially serves as a full census of reported outbreaks rather than a traditional sample, as all institutions are legally obligated to report under HPPA [@OntarioHealthAct1990]. However, certain cases are excluded, such as mild infections managed outside healthcare settings or individuals unable to seek timely medical care, resulting in potential underrepresentation.

@@

Outbreaks are identified and reported by healthcare institutions based on clinical symptoms or laboratory-confirmed cases. These reports are then submitted to Toronto Public Health, with updates provided weekly to maintain real-time relevance [@OntarioMinistry2024]. However, smaller outbreaks or those diagnosed late may be underreported, leading to data gaps. Variability in reporting practices across institution types further complicates data consistency; for instance, long-term care homes may report more frequently due to heightened surveillance requirements, whereas transitional care centers might report less rigorously.

@@

The dataset adheres to rigorous reporting standards mandated by the HPPA [@OntarioHealthAct1990], ensuring uniformity in data collection across institutions. Outbreaks are defined using standardized criteria, such as an increase in infection rates above expected baselines within a specific ward or institution. This ensures consistency in reported data. Institutions work closely with Toronto Public Health (TPH) to verify and classify outbreaks, further enhancing the dataset's reliability.

@@

The dataset supports broader public health efforts to manage and mitigate healthcare-associated infections, aligning with the objectives outlined in the 2023-2024 Canadian Public Health Plan [@Canada2024]. The outbreak monitoring framework is consistent with international practices, such as the World Health Organization's guidelines for disease surveillance and transmission [@citewho]. By linking local data collection to global health standards, this dataset highlights the critical role of systematic outbreak reporting in improving public health outcomes and guiding evidence-based policy decisions.
